---
title: "Drug combination analysis"
output: html_document
date: "2025-04-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Drug combination analysis

Use this script to generate a bar plot and to calcute synergy scores.

Found more information in [User tutorial of the SynergyFinder Plus] (https://bioconductor.org/packages/devel/bioc/vignettes/synergyfinder/inst/doc/User_tutorual_of_the_SynergyFinder_plus.html) by Zheng S et al.

First install and load the required packages.

```{r}
# To install pacman if needed
if (!require("pacman")) install.packages("pacman")

# Load requiered libraries
pacman::p_load(dplyr, ggplot2, synergyfinder, openxlsx)
```

## Load your data

You can also embed plots, for example:

```{r}
# set the working directory. If needed, uncomment the next line
#setwd("/path/to/your/directory")

# load the data
df <- read.csv("df_plot.csv") #to generate bar plot
sf_df <- read.csv("synergyfinder_df.csv") #to use synergy finder
```

## Bar plot with error bars

```{r}
# to set the bars order
label_order <- c(
  "NC", 
  "Dip4", 
  "Dip8", 
  "Dip16",
  "T-90",
  "T-180",
  "T-360", 
  "Dip4/T-90",
  "Dip4/T-180",
  "Dip4/T-360",
  "Dip8/T-90",
  "Dip8/T-180", 
  "Dip8/T-360", 
  "Dip16/T-90",
  "Dip16/T-180",
  "Dip16/T-360"
)

df$treatment <- factor(df$treatment, levels = label_order)

# Bar plot
pdf("ED_diptuc_rep1.pdf", width = 15, height = 10) #to export the plot in pdf format

ggplot(df, aes(x = treatment, y = mean)) +
  geom_bar(stat = "identity", fill = "skyblue", width = 0.7) +
  geom_errorbar(
    aes(ymin = mean - SD, ymax = mean + SD),  
    width = 0.2,
    color = "black",
    linewidth = 0.5
  ) +
  labs(
    x = "Treatment", 
    y = "Cell viability (%)", 
    title = "Cell viability by Dose (Mean ± SD)"
  ) +
  theme_bw()  

dev.off()

```

## Reshaping and pre-processing
The output data is a list containing following components:

drug_pairs: contains p-values for response value (adjusted and original input value).
response: contains the % of inhibition.The response values are the mean values over the replicates.contains the statistics (mean, standard error, 95% confidence interval) for the observed response. The “n” column records the number of replicate for each treatment.

```{r}
# Reshaphing
res <- ReshapeData(
  data = sf_df,
  data_type = "viability",
  impute = TRUE,
  impute_method = NULL,
  noise = TRUE,
  iteration = 10, #for treatments with replicates
  seed = 1)

# Export dataframes

response <- res$response
response_statistics <- res$response_statistics

write.xlsx(response, file = "reponse.xlsx")
write.xlsx(response_statistics, file = "response statistics.xlsx")

```

## Drug synergy scoring
The output is:

drug_pair: Contains the p-values for synergy scores.
synergy_scores: calculated from bootstrap samples
synergy_scores_statistics: containing the statistics (mean value, standard error of mean and bounders for 95% confidence interval).

```{r}
res <- CalculateSynergy(
  data = res,
  method = c("ZIP", "HSA", "Bliss", "Loewe"),
  Emin = NA,
  Emax = NA,
  correct_baseline = "non")

drug_pairs <- res$drug_pairs
synergy_scores <- res$synergy_scores
ss_statistics <- res$synergy_scores_statistics

# Export dataframes
write.xlsx(drug_pairs, file = "drug pairs.xlsx")
write.xlsx(synergy_scores, file = "synergy scores.xlsx")
write.xlsx(ss_statistics, file = "ss statistics.xlsx")
```

## Sensitivity scoring
The sensitivity scores are calculated from bootstrap samples 

```{r}
res <- CalculateSensitivity(
  data = res,
  correct_baseline = "non",
  iteration = 10 # Number of iterations for bootstrapping
)


sensitivity_statistics <- res$sensitivity_scores_statistics

#Export dataframe
write.xlsx(sensitivity_statistics, file = "sensitivity statistics.xlsx")
```

## Visualization

```{r}

# Standard error of mean response

pdf("Response heatmap.pdf", width = 15, height = 15)

Plot2DrugHeatmap(
    data = res,
    plot_block = 1,
    drugs = c(1, 2),
    plot_value = "response",
    dynamic = FALSE,
    statistic = "sem",
    summary_statistic = c("mean",  "median")
  )

dev.off()

# 95% confidence interval of Bliss synergy scores

pdf("Bliss synergy heatmap.pdf", width = 15, height = 15)

Plot2DrugHeatmap(
    data = res,
    plot_block = 1,
    drugs = c(1, 2),
    plot_value = "Bliss_synergy",
    statistic = "ci",
    dynamic = FALSE,
    summary_statistic = c("quantile_25", "quantile_75")
)

dev.off()

```


